Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-19 2017-Xudong-SMT/FM-2018-extended.tex
Minor modifications -- 15 pages now

\newcommand{\noteSB}[2][color=green!40, size=\tiny]{\todo[#1]{{\bf
      Note: } {#2}}}
\newcommand{\noteEM}[2][color=blue!40, size=\tiny]{\todo[#1]{{\bf Eric: } {#2}}}
\newcommand{\noteInSB}[2][inline,color=green!40]{\todo[#1]{{\bf Simon: } {#2}}}

@@

\item[$\bullet$] $\symb{SV}_k^{k\in K}$, with $K\in\I_\P$, is a set of
  synchronization vectors: for each $k\!\in\! K$,
  $\symb{SV}_k\!=\!\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k$ where
  $\alpha'_k\in \mathcal{A}_\P$, $I_k\subseteq I$, $J_k\subseteq J$,
  $\forall i\!\in\! I_k.\,\alpha_{i}\!\in\!\Sort(\pNet_i)$\noteSB{What is the sort of a pNet? Should we define $\Sort(\pNet_i) \triangleq \{\Label(\symb{SV}_k\,|\,k \in K\}$?}, 
  $\forall j\!\in\!
  J_k.\,\alpha_{j}\!\in\!S_j$, and $\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
  J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
$\Label(\symb{SV}_k) = \alpha'_k$.

@@

In appendix \ref{appendix:mapsmt} we formally define the translation of each
open transition, including:
- collect all variables in the transition, and declare them,
- check well-formness and correct typing of expressions
- translate the predicate into a conjunction of assertions
- if present in the source state, translate the state-variable
assignments into a disjunctive assertion.
Here again, this translation ensures that no runtime error will occur
when running the SMT engine.

In Fig.~\ref{schema:smt-lib} you can see the decomposition of the
predicate into a set of asserts, each encoding an elementary equality,
inequality, or a guard. The final command in the translation is a
\texttt{(check-sat)}, which result (sat or unsat) will be decoded by
our algorithm.

@@

Among the questions that have arised during this work, an important one is the
efficiency of Z3 for our needs. Z3 is famous for being very fast in
solving very large sets of assertions.%% , which could be important for
%% us
However, we encountered problems during the implementation of the
checking method using Z3. For example, if the $\mathit{functions}$ and the
$\mathit{datatypes}$ part of the algebra presentation that we submit to the solver
are independent from each other, then checking the $\mathit{assertions}$ is 
simple. However, $\mathit{functions}$ on recursive $\mathit{datatypes}$ make it
more complex. Special rules might be defined by the user for the
induction. Also, we need more evidence that dealing with more complex
action algebras, which would involve axiomatizing their data structures
into Z3 theories, will indeed allow us to decide validity of
predicates.
We may also try other automatic theorem provers, depending on the structure of the proofs we need.

\TODO{Change perspective: we are working on the formal extension of
  BIB architectures, and their semantics in terms of pNets. Then
  bisimulation and model-checking principles adapted to open pNet
  systems are under development}

Naturally, our next goals after the generation of the open automata
will be to model-check, and to check equivalence of  pNets. While
model-checking open automata (with a suitable logic including
predicates on data) seems easy to define, equivalence checking is more challenging.
In a previous paper~\hl{\cite{x}}, we
have already found the FH-bisimulation, inspired by~\cite{deSimone85}, to be a
suitable definition. However, the usual approach of computing bisimulation by partition
refinement does not work easily with symbolic transitions. In a first
step we may want instead to define and implement an algorithm for
checking that a given equivalence relation between open automata 
states is indeed a FH-Bisimulation. Finding automatically a
bisimulation relation or the most general bisimulation will be more
challenging. It appears also that when comparing different pNets with
different structure, strong bisimulation will be too strong a notion,
and that we will have to define proper notions of weak equivalence or
refinements.

>>>>

\newcommand{\noteSB}[2][color=green!40, size=\tiny]{\todo[#1]{{\bf
      Note: } {#2}}}
\newcommand{\noteEM}[2][color=blue!40, size=\tiny]{\todo[#1]{{\bf Eric: } {#2}}}
\newcommand{\noteInSB}[2][inline,color=green!40]{\todo[#1]{{\bf Simon: } {#2}}}
\newcommand{\todoMargin}[2][color=red!60, size=\tiny]{\todo[#1]{{\bf
      ToDo: } {#2}}}

@@

\item[$\bullet$] $\symb{SV}_k^{k\in K}$, with $K\in\I_\P$, is a set of
  synchronization vectors: for each $k\!\in\! K$,
  $\symb{SV}_k\!=\!\alpha_{l}^{l\in I_k \uplus J_k}\to\alpha'_k$ where
  $\alpha'_k\in \mathcal{A}_\P$, $I_k\subseteq I$, $J_k\subseteq J$,
  $\forall i\!\in\! I_k.\,\alpha_{i}\!\in\!\Sort(\pNet_i)$\noteSB{What is the sort of a pNet? Should we define $\Sort(\pNet) \triangleq \{\Label(\symb{SV}_k\,|\,k \in K\}$?}, 
  $\forall j\!\in\!
  J_k.\,\alpha_{j}\!\in\!S_j$, and $\vars(\alpha'_k)\subseteq \bigcup_{l\in I_k\uplus 
  J_k}{\vars({\alpha_l})}$. The global action of a vector $\symb{SV}_k$ is
$\Label(\symb{SV}_k) = \alpha'_k$.

@@

In App.~\ref{appendix:mapsmt} we formally define the translation of each
open transition, including:
- collect all variables in the transition, and declare them,
- check well-formness and correct typing of expressions
- translate the predicate into a conjunction of assertions
- if present in the source state, translate the state-variable
assignments into a disjunctive assertion.
Here again, this translation ensures that no runtime error will occur
when running the SMT engine.

Figure~\ref{schema:smt-lib} shows the decomposition of the
predicate into a set of asserts, each encoding an elementary equality,
inequality, or a guard. 
The result (sat or unsat) of the final command in the translation (\texttt{check-sat}) is decoded by
our algorithm.

@@

Among the questions that have arised during this work, an important one is the
efficiency of Z3 for our needs. Z3 is famous for being very fast in
solving very large sets of assertions. %% , which could be important for
%% us
However, we encountered problems during the implementation of the
checking method. %% using Z3. 
For example, if the $\mathit{functions}$ and the
$\mathit{datatypes}$ part of the algebra presentation that we submit to the solver
are independent from each other, then checking the $\mathit{assertions}$ is 
simple. However, $\mathit{functions}$ on recursive $\mathit{datatypes}$ make it
more complex. Special rules might be defined by the user for the
induction.  We need more evidence that dealing with more complex
action algebras, which would involve axiomatizing their data structures
as Z3 theories, will allow us to decide validity of
predicates.
We may also try other automatic theorem provers, depending on the structure of the proofs we need.

\todoMargin{Change perspective: we are working on the formal extension of
  BIP architectures, and their semantics in terms of pNets. Then
  bisimulation and model-checking principles adapted to open pNet
  systems are under development}
%
Naturally, our next goals after the generation of the open automata
will be to model-check, and to check equivalence of  pNets. While
model-checking open automata (with a suitable logic including
predicates on data) seems easy to define, equivalence checking is more challenging.
In a previous paper~\hl{\cite{x}}, we
have already found the FH-bisimulation, inspired by~\cite{deSimone85}, to be a
suitable definition. However, the usual approach of computing bisimulation by partition
refinement does not work easily with symbolic transitions. In a first
step we may want instead to define and implement an algorithm for
checking that a given equivalence relation between open automata 
states is indeed a FH-Bisimulation. Finding automatically a
bisimulation relation or the most general bisimulation will be more
challenging. It appears also that when comparing different pNets with
different structure, strong bisimulation will be too strong a notion,
and that we will have to define proper notions of weak equivalence or
refinements.
